---
title: "CO2 Emissions 1997"
author: "Team 2"
date: "2024-03-18"
output: 'pdf_document'  
---

```{r,echo=FALSE}
install.packages("blsR")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(magrittr)
library(patchwork)

library(readr)
library(dplyr)
library(ggplot2)

library(lubridate)

library(tsibble)
library(feasts)
#library(forecast)

library(sandwich)
library(lmtest)

library(nycflights13)
library(blsR)

library(Matrix) 
library(data.table) 
library(stats)
library(fable)
library(tseries)
library(reprex)
library(stargazer)
```

\centerline{\textit{POV: 1997}}

## Context 

Climate science has emerged as a leading field of interest in the 20th century. Efforts to bring awareness to the impact of human intervention on the environment, such as the burning of fossil fuels, have proved fruitful. The Intergovernmental Panel on Climate Change (IPCC) has reinforced these efforts. In 1990 they released the First Climate Assessment Report stating that "human activities are substantially increasing the atmospheric concentration of greenhouse gases" (IPCC, 1990). Greenhouse gases are a group of gases, such as carbon dioxide, methane, and nitrous oxide, that when present in higher concentrations in the atmosphere, raise the surface temperate of the Earth. Carbon dioxide is the most abundant greenhouse gas that is produced from human activity, namely energy production via fossil fuel combustion. Since the industrial revolution, energy consumption from petroleum and natural gas sources has risen dramatically. This report aims to investigate the following questions: How have the levels of atmospheric CO2 changed over time? And, is there an identifiable pattern that will persist into the future? Forecasting atmospheric carbon dioxide concentrations allows scientists to measure the corresponding impact to the global environment and justify the need for human intervention in the opposite direction. 

## Data and Exploration 

```{r load and format data, echo=FALSE}
data <- datasets::co2

data <- data %>% 
  as_tsibble(data) %>% 
  mutate(month = as.factor(month(index)))
```

Charles Keeling was a research scientist who made it his life's work to survey the atmosphere in hopes of confirming Svante Arrhenius's theory that fossil fuel combustion is increasing the concentration of CO2 in the atmosphere. To this end, Keeling collected atmospheric CO2 concentration measurements at a number of sampling-stations including the Mauna Loa Observatory in Hawaii. These measurements were taken using a CO2 analyzer which detects the amount of infrared absorption present in a air sample and turns it into a mole fraction of CO2, defined as the total CO2 molecules divided by the total non-water vapor molecules in the air, measured in parts per million (ppm). This report uses the data collected at the Mauna Loa Observatory between January 1959 and December 1997. The dataset consists of 468 observations with each observation representing the monthly total atmospheric concentration of CO2 (ppm). The observations for February, March, and April 1964 were unavailable so the values in the dataset were generated via linear interpolation between the observations for January and May 1964. 

In order to better understand the characteristics of this time series, we conducted a exploratory analysis prior to modeling. Figure 1 shows the time series plot for CO2 concentration, its autocorrelation plot, and its partial autocorrelation plot. The time series plot shows a clear positive trend as well as the presence of seasonality. The autocorrelation plot provides evidence to support the presence of both trend and seasonality as it decays with increasing lags and shows a spike at about every twelfth lag, indicating a seasonal cycle. 

- Discuss seasonal/trend decomposition 
- What about growth rates?

```{r time series plot, echo=FALSE, fig.width = 10, fig.height = 3}
time_series <- ggplot(data, aes(x=index, y=value)) +
  geom_line() +
  ylab("CO2 Concentration") +
  xlab("Time") +
  ggtitle("CO2 Concentration 1959 - 1997")

# acf <- ggAcf(data$value, lag.max = 50) +
#   ggtitle("ACF")
# 
# pacf <- ggPacf(data$value, lag.max = 50) +
#   ggtitle("Partial ACF")
# 
# time_series + acf + pacf + plot_annotation(expression(italic('Figure 1. Time Series, ACF, and PACF')), theme=theme(plot.title=element_text(hjust=0.5)))
```

## Models and Forecasts

- Why modeling is important to aid understanding? 
- Empirical evidence 

### *Linear Model*

```{r  linear model, echo = FALSE, results='hide'}
data$month_since_start <- as.numeric(index(data) - min(index(data))) + 1

lm_model <- lm(value ~ month_since_start, data=data)
summary(lm_model)
```
```{r  quadratic model, echo = FALSE, results='hide'}
quad_model <- lm(value ~ I(month_since_start^2) , data=data)
summary(quad_model)
```
```{r residual plots linear models, echo = FALSE}
lm_residual_data <- data.frame(
  predicted_values = fitted(lm_model),
  residuals = rstandard(lm_model)
)

quad_residual_data <- data.frame(
  predicted_values = fitted(quad_model),
  residuals = rstandard(quad_model)
)

lm_resid_plot <- ggplot(lm_residual_data, aes(x = predicted_values, y = residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_smooth(se = FALSE) +
  labs(title = "Linear Model Standardized Residuals vs Fitted Values", 
       x = "Fitted Values", y = "Standardized Residuals")

quad_resid_plot <- ggplot(quad_residual_data, aes(x = predicted_values, y = residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_smooth(se = FALSE) +
  labs(title = "Quadratic Model Standardized Residuals vs Fitted Values", 
       x = "Fitted Values", y = "Standardized Residuals")

lm_resid_plot / quad_resid_plot
```
```{r polynomial time trend model, echo = FALSE}
lm_seasonal_model <- lm(value ~ month_since_start + month, data=data)
summary(lm_seasonal_model)
```

- Functional form: y = t + e (co2 concentration = time index + error) &  y = t^2 + e
- Results
  - High R-squared values 
  - Significant p-values 
- Interpretation/evaluation? 
  - U-shaped (and upside down U-shaped) patterns
  - Suggest linear models are not appropriate for the data
  - Linear model predicting too low middle values 
  - Quadratic model predicting too high middle values 

### *ARIMA Model*

```{r ARIMA Modeling, echo = FALSE}
model_ma.bic<-data %>%
model(ARIMA(value ~ 1 + pdq(0,0:1,0:3) + PDQ(0,0:1,0:3), ic="bic", stepwise=F, greedy=F,
            order_constraint = p + q + P + Q <= 10))

model_ar.bic<-data %>%
model(ARIMA(value ~ 1 + pdq(0:3,0:1,0) + PDQ(0:3,0:1,0), ic="bic", stepwise=F, greedy=F,
            order_constraint = p + q + P + Q <= 10))

model_full.bic<-data %>%
model(ARIMA(value ~ 1 + pdq(0:3,0:1,0:3) + PDQ(0:3,0:1,0:3), ic="bic", stepwise=F, greedy=F,
            order_constraint = p + q + P + Q <= 10))

# Report model summaries 
model_ma.bic %>% # Best BIC model to use for forecasting 
report()

model_ar.bic %>%
report()

model_full.bic %>%
report()

```

- Functional form 
- Results 

### *Forecasts*

```{r polynomial forecasts, echo = FALSE, results = 'hide'}
library(forecast)
index <- seq(from = ymd("1998-01-01"), to = ymd("2020-12-01"), by = "1 month")

ext_data <- data.frame(index = index) %>% 
  as_tsibble(index=index) %>% 
  mutate(month = as.factor(month(index))) 

ext_data$month_since_start <- seq(max(data$month_since_start)+1,max(data$month_since_start)+(23*12))

forecast <- data.frame(predict = predict(lm_seasonal_model, newdata = ext_data))

forecast$index <- ext_data$index

ggplot(forecast, aes(x=index, y=predict)) +
  geom_line() +
  ylab("CO2 Concentration") +
  xlab("Year-Month") +
  ggtitle("CO2 Concentration Forecast 1998 - 2020")
```

```{r ARIMA forcast 2022, echo=FALSE}
model_ma.forecasts<-forecast(model_ma.bic, h=25*12)

autoplot(model_ma.forecasts) +
  autolayer(data, colour = "black", .vars = value) +
  xlab("Year") +
  ylab("Value") +
  ggtitle("ARIMA Forecast for Next 25 Years")
```

- Forecasts 
- Predictions for when CO2 is expected to be at 420 ppm and 500 ppm 
- Interpretation/evaluation? 


#(1 point) Task 0b: Introduction

In this introduction, you can assume that your reader will have just read your 1997 report. In this introduction,
very briefly pose the question that you are evaluating, and describe what (if anything) has changed in
the data generating process between 1997 and the present.


#(3 points) Task 1b: Create a modern data pipeline for Mona Loa CO2 data.

The most current data is provided by the United States’ National Oceanic and Atmospheric Administration,
on a data page [here]. Gather the most recent weekly data from this page. (A group that is interested in
even more data management might choose to work with the hourly data.)
Create a data pipeline that starts by reading from the appropriate URL, and ends by saving an object called
co2_present that is a suitable time series object.
Conduct the same EDA on this data. Describe how the Keeling Curve evolved from 1997 to the present,
noting where the series seems to be following similar trends to the series that you “evaluated in 1997” and
where the series seems to be following different trends. This EDA can use the same, or very similar tools
and views as you provided in your 1997 report.

```{r}
weekly_co2_url <- "https://gml.noaa.gov/webdata/ccgg/trends/co2/co2_weekly_mlo.csv"

content <- read_lines(weekly_co2_url, skip_empty_rows = TRUE)

header_end_index <- max(grep("^#", content))

co2_present <- read_csv(weekly_co2_url, skip = header_end_index, show_col_types = FALSE)

co2_present <- co2_present %>%
  mutate(date = make_date(year, month, day))

co2_present <- as_tsibble(co2_present, index = date)

co2_present <- co2_present[co2_present$average != -999.99, ]

head(co2_present)
```
```{r}
dim(co2_present)
```

```{r}
ggplot(co2_present, aes(x = date, y = average)) + 
  geom_line() + 
  theme_minimal() +
  labs(title = "Time Series of CO2 Levels", x = "Date", y = "CO2 Concentration")
```
The plot shows a clear upward trend, indicating that CO2 levels have been increasing over the given time period. The pattern also shows seasonal fluctuations within each year, where the CO2 concentration peaks and then drops slightly, before rising again. The overall trend, however, is an increase in CO2 levels.

```{r}
acf(co2_present$average, na.action = na.pass)

```
The plot shows a strong positive autocorrelation at all lags up to 35, and all are above the significance level, indicating a very persistent time series with a strong seasonal or cyclic pattern. This might suggest that the time series data of CO2 concentrations have a consistent pattern that repeats over time, with no significant decay in correlation as the lag increases.

```{r}
pacf(co2_present$average, na.action = na.pass)

```
The plot indicates that there is no almost no significant partial autocorrelation in the data at lags greater than zero. This could suggest that a simple autoregressive model may not be a good fit for the data.

```{r}
ggplot(co2_present, aes(x = average)) + 
  geom_histogram(binwidth = 1, fill = 'blue', color = 'black') +
  theme_minimal() +
  labs(title = "Histogram of CO2 Levels", x = "CO2 Concentration", y = "Frequency")
```
The plot shows multuple peaks suggesting more of a multimodal distribution. This could imply that there are multiple common CO2 levels within the data, possibly reflecting different environmental conditions or measurement periods. The data appears to be right-skewed. 


#(1 point) Task 2b: Compare linear model forecasts against realized CO2

Descriptively compare realized atmospheric CO2 levels to those predicted by your forecast from a linear time
model in 1997 (i.e. “Task 2a”). (You do not need to run any formal tests for this task.)


```{r month data}

co2_present <- read_csv(weekly_co2_url, skip = header_end_index, show_col_types = FALSE)

co2_present <- co2_present[co2_present$average != -999.99, ]

co2_monthly <- co2_present %>%
  group_by(year, month) %>%
  summarise(average = mean(average, na.rm = TRUE),
            .groups = "drop")

co2_monthly <- co2_monthly %>%
  mutate(Month = make_date(year, month, day = 1)) %>%
  select(-year, -month)  

co2_monthly_present <- as_tsibble(co2_monthly, index = Month)

head(co2_monthly_present)
```

#(1 point) Task 3b: Compare ARIMA models forecasts against realized CO2
Descriptively compare realized atmospheric CO2 levels to those predicted by your forecast from the ARIMA
model that you fitted in 1997 (i.e. “Task 3a”). Describe how the Keeling Curve evolved from 1997 to the
present.

#(3 points) Task 4b: Evaluate the performance of 1997 linear and ARIMA models
In 1997 you made predictions about the first time that CO2 would cross 420 ppm. How close were your
models to the truth?
After reflecting on your performance on this threshold-prediction task, continue to use the weekly data to
generate a month-average series from 1997 to the present, and compare the overall forecasting performance
of your models from Parts 2a and 3b over the entire period. (You should conduct formal tests for this task.)


#(4 points) Task 5b: Train best models on present data
Seasonally adjust the weekly NOAA data, and split both seasonally-adjusted (SA) and non-seasonallyadjusted
(NSA) series into training and test sets, using the last two years of observations as the test sets.
For both SA and NSA series, fit ARIMA models using all appropriate steps. Measure and discuss how your
models perform in-sample and (psuedo-) out-of-sample, comparing candidate models and explaining your
choice. In addition, fit a polynomial time-trend model to the seasonally-adjusted series and compare its
performance to that of your ARIMA model.

#(3 points) Task Part 6b: How bad could it get?
With the non-seasonally adjusted data series, generate predictions for when atmospheric CO2 is expected
to be at 420 ppm and 500 ppm levels for the first and final times (consider prediction intervals as well as
point estimates in your answer). Generate a prediction for atmospheric CO2 levels in the year 2122. How
confident are you that these will be accurate predictions?